---
title: "Impact of regulation changes and interest rates on private company acquisitions"
author: "Niklas Nihtila"
date: "4/25/2021"
output:
    html_notebook:
        toc: yes
        toc_float: yes
---
```{r, message = FALSE}
library(ISLR)
library(mosaic)
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(MASS)
library(gbm)
library(GGally)
library(broom)
library(class)
library(tree)
library(dplyr)
library(anytime)
library(xts)
library(lubridate)
library(caret)
```


# Abstract

Over the last two decades we have seen a decrease in Initial Public Offering  activity, and an increase in Private Equity activity, due to the passing of legislation that were favorable to the Private Equity industry and unfavorable to public markets and increased availability of capital within the private markets. This has helped fuel the private company acquisition market. 

We use company acquisition data from DealStats to assess the impact of different regulations and the interest rate environment on the private company acquisition prices between 1999 and 2018. The results of the analysis points that even though these regulation as well as interest rates could have had an impact on acquisition prices the main impact is from the overall economic environment and sentiment. We see high acquisition prices and activity in times of economic growth and slowdown in recessions and as regulations are usually enacted as response to economic crises and the interest rate environment is highly correlated with the economic the variables show unexpected relationships with acquisition prices.

# Introduction

Private company acquisitions have been common for decades. Acquisitions allow companies to grow faster and it helps them move into new markets quickly. One large contributor to the total private acquisition market is the Private Equity industry. Private Equity buyout funds, which became popular in the 1980's, complete vast amounts of acquisitions of both private and public companies in the hopes of streamlining the companies for higher efficiency and high returns. The rise of private equity in the last few decades has been helped by increasing capital availability and changes in the regulatory environment. These changes have also affected the public equity markets significantly as there has been a sharp decline in Initial Public Offerings. 

The aim of this study is to look at how these regulation environment changes as well as the interest rate environment has affected the size of all private company acquisitions. The regulations that the study takes into account are the Regulation Fair Disclosure, Sarbanes-Oxley Act, Global Analyst Research Settlement and the Jumpstart our Business Start-Ups act. The other important portion is the effect of the interest rate environment on the acquisition prices which will be assessed by looking at the impact of changes in the federal funds rate. Even though the effect of the regulation pieces have been studied in-depth, the effect of the pieces to the overall acquisition market has not been looked at before.

Previous work has found the three of the four regulation changes to have affected mainly the IPO market. Sarbanes-Oxley act and the Global Analyst Research Settlement contributed to the price of either going public or staying public. Sarbanes-Oxley act of 2002 is found to have increased compliance costs for public companies with under $1 billion in revenue by 171% in average. The Global Analyst Research Settlement of 2003 affected IPO costs directly, one study found that following the Global Settlement it has become more difficult to secure equity research for IPOs and the costs of that research have become more expensive.

Regulation Fair Disclosure's affect on public and private markets is expected to be minimal. The main impact is by increased liability and slight increase compliance costs as public companies are required to disclose any material information to the public.

The JOBS act  of 2012 affected both public and private markets, through increased capital availability in the private markets and looser IPO regulation for smaller companies. The main idea behind the JOBS act was to increase IPO activity by allowing small-cap companies to test the waters for an IPO and disclose less information to the public. However with allowing easier ways to raise capital private companies the impact was on the contrary, companies stayed private longer and enjoyed the privacy of not being a public company.

All these regulations were not built to have direct impact on the private company acquisitions. However, as less companies set their sights on an IPO other exit strategies became more popular, especially in the Private Equity industry. Companies were more inclined to be acquired than to go public as the premium for acquisitions lowered due to the increased costs of going public. This is expected to translate to increase in acquisition activity as well as increase in acquisition prices.

Unlike the regulations above, the interest rate environment has a direct impact on company valuations. Not only do higher interest rates lower the incentive of leveraged buy-outs but they also affect company valuations negatively. The valuation industry relies heavily on discounted cash flow models, expected future cash flows are discounted at a discount rate (higher discount rate equals lower valuation). The discount rate in turn takes a risk-free rate, usually the yield of a 10-yr or 20-yr treasury bill, into account and as the risk-free rate rises so does the overall discount rate. 



# Methods

Maybe some short intro


### Importing Datasets

```{r}
Deal_Stats <- read.csv(file = "FINC-412DS-MasterData3-20-2019--2.0(1).csv", header=TRUE, sep=",")
int_rate <- read.csv(file = "int_rate.csv", header = TRUE, sep = ",")
CPI <- read.csv(file = "CPI_U_BLS.csv", header = TRUE, sep = ",")
```

The main dataset used in this paper is a datapull from DealStats. Dealstats is a database containing historical company acquisitions and is mostly used for acquisition valuation purposes. In this case the data pulled from DealStats contains `r nrow(Deal_Stats)` acquistions over a time period from 1999 to 2018. The dataset includes acquisitions done outside the US as well as public company acquistions. As the purpose of the study is to look at private company transactions in the US I have filtered out acquisitions of public companies, but the dataset still includes acquisitions of private companies by public entities as these are still categorizes as private company acquisitions. In addition acquisitions where both parties are foreign companies have been filtered out to only focus on transactions of US based companies.

```{r}
Deal_Stats <- filter(Deal_Stats, Deal_Stats$TargetType != "Public")
Deal_Stats <- filter(Deal_Stats, Deal_Stats$TargetCountry == "United States" | Deal_Stats$AcquirerCountry == "United States")
```

Every acquisition entry also has the percentage acquired by the owner. In almost all acquisitions the percentage acquired is 100% but we do have some NA values. Due to not knowing the percentage acquired the MVICPrice might distort the actual transaction and thus the NA values will be filtered out of the dataset.

```{r}
Deal_Stats$PercentageAcquired <-  as.numeric(sub("%","",Deal_Stats$PercentageAcquired))/100
Deal_Stats %>% count(PercentageAcquired != 1)
```

```{r}
Deal_Stats <- filter(Deal_Stats, Deal_Stats$PercentageAcquired == 1)
```

The remaining number of acquisitions is `r nrow(Deal_Stats)`.

### Cleaning CPI dataset (https://stackoverflow.com/questions/50167509/converting-monthly-data-to-daily-in-r)

As the transactions span over a 20 year period, the impact of inflation has to be taken into account. To achieve this I will add a variable CPI_U which represents the monthly CPI value for all urban consumers and for all items. I believe this to be the most accurate representation of the inflation during the 20 year period that is being analyzed in this paper, due to the fact that the dataset contains a very diverse set of transactions in a very diverse set of industries.

```{r}
CPI$DATE <- as.Date(paste(CPI$Year,CPI$Month,"01", sep = "/"))
CPI <- do.call("rbind", lapply(1:nrow(CPI), function(i) 
data.frame(DATE = seq(CPI$DATE[i], 
                  (seq(CPI$DATE[i],length=2,by="months") - 1)[2], by = "1 days"), 
                  CPI_U = CPI$CPI_U[i])))
```

### Creating Regulation Variables

In this study we focus on four regulation pieces that have been enacted during the timeline of the dataset, between 1999 to 2018. The four pieces in question are Regulation FD (Reg_FD) , Sarbanes-Oxley act (SOX), Global Analyst Research Settlement (GARS) and Jumpstart our Business Start-Ups act (JOBS). These four regulation pieces will represented by factor variables with labels Yes and No that are distinguished by the date when they were enacted.

```{r}
Reg_FD <- anydate("Oct 23 2000")
SOX <- anydate("Jul 30 2002")
GARS <- anydate("Apr 28 2003")
JOBS <- anydate("Apr 5 2012")
```


### Interest rate

In addition to the regulation variables and the CPI variable another point of interest is the interest rate environment. For this research we are using the federal funds rate (FFR) to represent the overall environment. The federal funds rate is the main tool that the federal reserve or the federal open market committee uses to impact the interest rate environment. It is the base rate at which commercial banks loan overnight deposits to each other to meet the required percentage of deposits in an account at the federal reserve bank. The reason why we use this in the study is due to its impact on the overall credit environment as it indirectly affects other borrowing rates. However, it is important to notice that it has a more significant effect on other rates, such as short-term lending and a less significant or no effect on more risky loans such as junk bonds. 

### Merging Datasets

To perform the data analysis the datasets and all required variables are merged into a single dataset.

```{r}
Deal_Stats <- transform(Deal_Stats, SaleDate = anydate(SaleDate))
int_rate <- transform(int_rate, DATE = anydate(DATE))
Data <- merge(x = Deal_Stats, y = int_rate, by.x = "SaleDate", by.y = "DATE")
Data <- merge(x = Data, y = CPI, by.x = "SaleDate", by.y = "DATE") 
Data <- Data %>%
  mutate(Reg_FD = case_when(SaleDate >= Reg_FD ~ 1, SaleDate < Reg_FD ~ 0))
Data <- Data %>%
  mutate(SOX = case_when(SaleDate >= SOX ~ 1, SaleDate < SOX ~ 0))
Data <- Data %>%
  mutate(GARS = case_when(SaleDate >= GARS ~ 1, SaleDate < GARS ~ 0)) 
Data <- Data %>%
  mutate(JOBS = case_when(SaleDate >= JOBS ~ 1, SaleDate < JOBS ~ 0))
Data$Reg_FD <-  factor(Data$Reg_FD, levels = c(0,1), labels = c("No","Yes"))
Data$SOX <-  factor(Data$SOX, levels = c(0,1), labels = c("No","Yes"))
Data$GARS <-  factor(Data$GARS, levels = c(0,1), labels = c("No","Yes"))
Data$JOBS <-  factor(Data$JOBS, levels = c(0,1), labels = c("No","Yes"))
```


### Filtering out unnecessary variables

The complete dataset contains 170 variables including the ones added for the purpose of this paper. In this analysis we will only make use of 14 variables; PercentageAcquired, MVICPrice, Sale date, Feds Fund rate, regulation variables, Net Sales and CPI_U. The other variables that are not the main focus of this paper are used for the following purposes. PercentageAcquired is used to assess the real amount of transaction as the MVICPrice, or market value of invested capital, represents the enterprise value of whole the entity and Net sales variable is used to capture the variation between the size difference of the acquired companies. These variables will be added into a new dataset called Df which will be used for the rest of the study.

```{r}
Df <- data.frame("PercentageAcquired"= Data$PercentageAcquired, "MVICPrice" = Data$MVICPrice, "SaleDate" = Data$SaleDate, "FFR" = Data$DFF, "Reg_FD" = Data$Reg_FD, "SOX" = Data$SOX, "GARS" = Data$GARS, "JOBS" = Data$JOBS, "NetSales" = Data$NetSales, "EBIT" = Data$TargetEBIT, "CPI_U" = Data$CPI_U)
Df$MVICPrice <- as.numeric(gsub('[$,]', '', Df$MVICPrice))
Df$NetSales <- as.numeric(gsub('[$,]', '', Df$NetSales))
```


## Dataset Trends

To get a better grasp on the some crucial trends of the private company acquisition market as well as the macroeconomic variables I looked at trends in the acquisition number, sizes and in the federal funds rate.

### Number of transactions per Month


```{r}
Df <- separate(Df, "SaleDate", c("Year", "Month", "Day"), sep = "-")
Df$SaleDate<-as.Date(with(Df,paste(Year,Month,Day,sep="-")),"%Y-%m-%d")
Transaction_by_month <- Df %>% count(Year,Month)
Transaction_by_month$Date <- paste(Transaction_by_month$Year,Transaction_by_month$Month,"01", sep = "-")
Transaction_by_month$Date <- as.Date(Transaction_by_month$Date)
ggplot() + 
  geom_line(data = Transaction_by_month, aes(x = Date, y = n)) +
  labs(title = "Acquisition Transactions by Month",
y = "Number of Transactions") 
```

### Federal Funds Rate

```{r}
ggplot(Df , aes(x = SaleDate, y = FFR, group = 1)) +
  geom_line() +
  labs(title = "Federal Funds Rate", y = "Federal Funds Rate (%)", x = "Date")
```

### MVIC/Date

```{r}
TotalSize_Month <-  Df %>% 
    group_by(Year = year(SaleDate) ,Month = month(SaleDate,label = T)) %>% 
    summarise(MVICPrice = sum(MVICPrice, na.rm = T))
TotalSize_Month$Date <- paste(TotalSize_Month$Year,TotalSize_Month$Month,"01", sep = "-")
TotalSize_Month$Date <- as.Date(Transaction_by_month$Date)
ggplot() + 
 geom_line(data = TotalSize_Month, aes(x = Date, y = MVICPrice, group = 1)) +
  labs(title = "Acquisition Sizes by Month",
y = "Total Size of Transactions")
```


## Exploratory Data Analysis

Exploratory data analysis methods, graphing summarizing, was used to help distinguish the variable distributions and possible relationship as well as help determine the usability of different models.

### Summary

The summary below shows the distributions of all variables in the Df dataset that will be used in the modeling.

```{r}
summary(Df)
```

Based on the summary we have acquisitions ranging from zero dollars to over 17 billion dollars, similarly we have NetSales with a similar range from 0 dollars to over 17 billion dollars. The dataset includes acquisitions with the MVICPrice at 0 dollars, this is either due to a missing value for the MVICPrice or some other agreement with a 0 dollar acquisition price. These acquisitions do not have any impact in the overall acquisition market size but do impose outliers for the data analysis which is why they will be filtered from the dataset.

```{r}
Df <- filter(Df, Df$MVICPrice != 0)
Df <- filter(Df, Df$NetSales > 0)
summary(Df)
```
The distributions of these two variables are quite similar with some differences. The NetSales mean is slightly lower than the mean MVICPrice, but the 1st and 2nd quartiles are higher for NetSales. For the Regulations we see the number of acquisition before and after each regulation. For Reg_FD, SOX and GARS the number of acquisitions before the regulation is much higher due to being enacted in the early 2000's and for JOBS there are more acquisitions before as the act was enacted in 2014. 

### MVICPrice and NetSales/CPI variables

To visualize the relationship between NetSales and MVICPrice we can plot a scatterplot:

```{r}
ggplot(Df, aes(x = MVICPrice, y = NetSales)) +
  geom_point(size = 0.2 , alpha = 0.2) +
  labs(title = "Net Sales to MVICPrice",
x = "NetSales", y = "MVICPrice")
```

Both variables are highly skewed to the right due to the large variance in acquisition price and target company sizes.
To get a better understanding of the relationship we can use a logarithmic transformation on both variables.

```{r}
ggplot(Df, aes(x = log(MVICPrice), y = log(NetSales))) +
  geom_point(size = 0.2, alpha = 0.2) +
  labs(title = "Log Net Sales to Log MVICPrice",
x = "Log(NetSales)", y = "Log(MVICPrice)")
```

With the logarithmic transformation we see a strong positive correlation between the two variables, which is as expected. The size of the company's revenue directly translates to the acquisition price.

To visualize the relationship between CPI and the log transformed MVICPrice we can plot a scatterplot:

```{r}
ggplot(Df, aes(x = CPI_U, y = log(MVICPrice ))) +
  geom_point(size = 0.3, alpha = 0.2)  +
  labs(title = "CPI to Log MVICPrice",
x = "CPI", y = "Log(MVICPrice)")
```

There doesn't seem to be a relationship between the two variables. The distribution of MVICPrice between the different CPI levels seems to be similar. We see slightly higher amounts of acquisitions with CPI between 210 and 240, which is due to a short deflation period and slower inflation after the financial crisis.

### MVICPrice and FFR/Regulation variables

Below is a scatterplot to visualize the relationship between the Federal Funds rate and the log transformed MVICPrice:

```{r}
ggplot(Df, aes(x = FFR, y = log(MVICPrice) )) +
  geom_point(size = 0.3, alpha = 0.2) +
  labs(title = "Federal Funds Rate to Log MVICPrice",
x = "Federal Funds Rate (%)", y = "Log(MVICPrice)")
```

The scatterplot shows little or no relationship between the two variables. The MVICPrice distribution seems relatively constant over different Federal Funds rates. However, we see a large cluster of acquisitions with the Federal Funds rate near zero. This is due to the lenghtened time the rate was kept low during the stagnation period after the financial crisis os 2008. The rate was kept low due to stagnation worries, low GDP growth, to stimulate the economy. The rate was kept around 0 percent from late 2008 to late 2015, which is approximately 35 percent of the whole time period of the dataset.

To visualize the distributions of the MVICPrice for the regulation variables I will use boxplots. Below is a boxplot visualizing the MVICPrice distribution before and after Regulation Fair Disclosure:

```{r}
ggplot(Df,  aes(y = log(MVICPrice), x = Reg_FD)) +
geom_boxplot() +
  labs(title = "Acquisition Prices Before and After Regulation FD",
x = "Regulation FD enacted")
```

There is visible difference between acquisition prices before and after the enactment of Regulation FD. The mean log MVICPrice is 2.5 higher before than after the enactment. After the regulation was enacted there is more variability in the values with a range from over 22.5 to almost 0. It is important to notice that the dataset only contains 1910 acquisitions before the regulation and over 25100 after as the regulation was enacted in october 2000.

Below is a boxplot between SOX and MVICPrice:

```{r}
ggplot(Df, aes(y = log(MVICPrice), x = SOX)) +
  geom_boxplot() +
  labs(title = "Acquisition Prices Before and After SOX",
x = "SOX enacted") 
```

The distribution before and after the enactment is quite similar to 

Similar to Regulation FD there is a visible difference, however not as significant as for Regulation FD. The difference seems similar as with Regulation FD, however the difference between the means is much smaller. As with Regulation FD it is important to notice that the SOX variable captures more acquisitions after the enactment, 3530 before and over 23530 after. 


Below is a boxplot between the Global Analyst Research Settlement and MVICPrice:

```{r}
ggplot(Df, aes(y = log(MVICPrice), x = GARS)) +
geom_boxplot() +
  labs(title = "Acquisition Prices Before and After the Global Analyst Research Settlement",
x = "Global Analyst Research Settlement enacted")
```

The Global Analyst Research Settlement has a similar MVICPrice distribution as Regulation FD and SOX, however the difference between before and after the enactment is becoming smaller. As all three regulations have a similar trend it is important to realize that this most likely due to the overlap these regulations have. All the regulation variables are a single point in time and look at the difference before and after. In the case of GARS, Reg FD and SOX the time difference is small so the distributions before and after do not differ much. This is due to all three regulation variables recording a large number of same transactions before and after.

Below is a boxplot between the Jumpstart our Business Start-Ups act and MVICPrice:

```{r}
ggplot(Df, aes(y = log(MVICPrice), x = JOBS)) +
geom_boxplot() +
  labs(title = "Acquisition Prices Before and After JOBS Act",
x = "JOBS Act enacted")
```

Unlike the other three regulations the JOBS act does not show a significant difference before and after the enactment. The distributions are very similar with some minor differences. 

With the preliminary data analysis from the explanatory variables that are of interest for this study, Regulation FD shows the largest difference in MVICPrice distributions. The JOBS act and the Federal Funds rate show little or no effect on the response variable MVICPrice.


### MVICPrice Histogram

```{r}
ggplot(Df, aes(x= log(MVICPrice))) +
  geom_histogram(binwidth = 0.5) +
  labs(title = "MVICPrice Distribution",
y = "Count", x = "Log(MVIC Price)")
```

### NetSales Histogram

```{r}
ggplot(Df, aes(x= log(NetSales))) +
  geom_histogram()
```
## Models

In this analysis we are looking at the dataset as a whole and focusing on the predictive power and significance of the interest rate and regulation variables. To assess this I will look at three different models, a multiple linear regression, Lasso regression and a Decision tree. Due to some specifics of the dataset which will be covered later the results will be taken from the multiple linear regression model.


## Multiple Linear Regression model

The multiple linear regression model used in this study has the following format:

$$\widehat{log(MVICPrice)} = \hat{\beta}_0 + \hat{\beta}_1 * {FFR} +\hat{\beta}_2 * {RegFD} + \hat{\beta}_3 * {SOX} + \hat{\beta}_4 *{GARS} + \hat{\beta}_5 *{JOBS} + \hat{\beta}_5 *{CPIU} + \hat{\beta}_6 *{log(NetSales)}  $$

We can run the model using the following code:

```{r}
model_lm <- lm(log(MVICPrice) ~ FFR + Reg_FD + SOX + GARS + JOBS + CPI_U + log(NetSales), data = Df)
```

To model results in the following statistics.

```{r}
glance(model_lm)
```

The model has an R squared of 0.785, however it is good to remember that we have a NetSales variable inputted in the model to control for variance. NetSales is highly correlated with the MVICPrice and is most likely the most predictive variable in the model and causes the r squared to increase significantly. The R squared for a model only using the variables of interest would be much lower as even though these variables might have an effect on the MVICPrice they show broader macroeconomic conditions and do not act as powerful predictors for single company transactions.

## Lasso

The second model I looked at was a lasso regression, which is similar to multiple linear regression with one significant difference. Lasso uses shrinkage to assess the impact of different variables and if the impact is insignificant, it disregards the variable. However there is one significant issue with Lasso regression in this case. We are multiple factorized categorical variables to assess the impact of the regulation pieces on the MVICPrice. With Lasso the results for these categorical variables might become distorted due to the mechanics behind Lasso which is why I am disregarding the results.




```{r}
grid <- 10^seq(2, -2, length = 100)
train_control_cv <- trainControl(method = "cv")
set.seed(1)
lasso_mod_caret <- train(log(MVICPrice) ~ FFR  + Reg_FD + SOX + GARS + JOBS + CPI_U + log(NetSales), data = Df,
                         trControl = train_control_cv,
                         preProcess = c("center", "scale"),
                         method = "glmnet",
                         tuneGrid = data.frame(alpha = 1, lambda = grid))
```

```{r}
best_lambda_lasso_caret <- lasso_mod_caret$bestTune$lambda
best_lambda_lasso_caret
```
```{r}
ggplot(lasso_mod_caret) +
    geom_vline(xintercept = best_lambda_lasso_caret, color = "red")
```
```{r}
coef(lasso_mod_caret$finalModel, s = best_lambda_lasso_caret)
```
```{r}
lasso_mod_caret
```

## Decision Tree

The final Model used in the analysis is a decision tree. A decision tree is a tree-like model that tries to predict a value based on different input variables. It works similar to a flow-chart diagram where nodes represent a classification output. As with Lasso the results are quite distorted due to the specifics of the dataset. Due to large variation between company acquisition we have take the variance into account to properly assess the impact of the variable that are of interest. In this case we use NetSales, however as acquisition prices are highly correlated with the target company sales the predictive power of this variable is much stronger than with any of the other variables, which poses a problem for the decision tree. As seen in the results below the decision tree only takes into account the NetSales figure for the analysis and disregards any other variables. This is why the model will not be used to assess the research question.

```{r}
train_rows_df <- createDataPartition(Df$MVICPrice, p = 0.5, list = FALSE)
train_df <- Df[train_rows_df, ]
test_df <- Df[-train_rows_df, ]
grid <- data.frame(cp = seq(from = 0.01, to = 0.2, by = 0.01))
sum(is.na(train_df$MVICPrice))
```


```{r}
fit <- train(log(MVICPrice) ~ FFR + Reg_FD + SOX + GARS + JOBS + log(NetSales) + CPI_U, data = train_df, tuneGrid = grid,
                    method = "rpart")
fit
```

```{r}
rpart.plot(fit$finalModel)
```

# Results

For the results we will use the results from the Multiple Linear Regression model. The variables we are most interested in are the FFR, which represents the federal funds rate and the regulation pieces which are represented by the following variables:

- Reg_FD (Regulation Fair Disclosure)
- Sox (Sarbanes-Oxley Act)
- GARS (Global Analyst Research Settlement)
- JOBS (Jumpstart our Business Start-Ups Act)

Below we can see the results of the model.

```{r}
summary(model_lm)
```
To address the skewdness in the MVICPrice and NetSales a log transformation was applied to both variables. This changes the interpretation of the coefficients. Instead of representing a unit change in the response variable the coefficients for variables that remain unchanged represent a % increase in MVICPrice when:

$$ \widehat{{\%} {\Delta} {MVICPrice}} = (e{^{Coefficient Estimate}}-1)*100$$
This changes the coefficients to the following:

```{r}
coefficients <-  (exp(model_lm$coefficients)-1)*100
coefficients[2:7]
```

These coefficients show the % change in MVICprice if everything else is constant. If we hold every other variable constant we can interpret them as follows:

A unit increase in the Federal Funds rate will increase the MVICPrice by 2.75%.

The enactment of Regulation FD resulted in a 43.72% decrease in the MVICPrice.

The enactment of the Sarbanes-Oxley acted resulted in a 9.96% decrease in the MVICPrice.

The enactment of the Global Analyst Research Settlement resulted in a 21.14% increase in the MVICPrice.

The enactment of the Jumpstart our Business Start-ups Act resulted in a 5.43% increase in the MVICPrice.

A unit increase in CPI_U will decrease the MVICPrice by 0.32%.


For the NetSales variable the interpretation is different as both the explanatory and the response variables are log transformed. In this case we do not modify the original coefficient, but we interpret it as when the explanatory variable increases by 1% the response variable changes by the coefficient % when everything else is held constant. For the NetSales the interpretation is as follows:

```{r}
model_lm$coefficients[8]
```
 
 A one percent increase in NetSales increases the MVICPrice by 1.05%.


# Discussion

The results are unexpected, in some cases the model points to a opposite relationship than what was expected. For the federal funds rate we have a positive relationship with the MVICPrice. However, when comparing to company valuation methods a higher risk-free interest rate translates to a higher cost of capital which decreases company valuations. The results seem to be contrary to what widely used methods would suggest. For the regulation variables the coefficients are surprisingly high. The model expects that Regulation Fair Disclosure, which affected only public companies, would decrease the MVICPrice by over 40%. Similarly the Global Analyst Research Settlement would increase the MVICPrice by over 15%.

Safe to say that the results seem to be pointing to another relationship than a direct causal relationship between the MVICPrice and the explanatory variables and possible issues with the analysis.

The main issue is that all variables used are tied to macroeconomic trends. As seen before the transaction prices and numbers follow closely the macroeconomic conditions at the time of the transaction. Now this is expected as when the economy grows you would expect more activity in the private markets as well as the public markets.
However, in addition to the transactions the explanatory variables are also tied to the overall economic conditions.
Regulations are usually enacted as a response for a crisis or some macroeconomic events, for example, 
SOX was enacted as a response to major financial scandals of public companies such as Enron, Tyco and WorldCom, and the Federal Funds rate acts as a response to macroeconomic trends.

When you have high growth in the economy, like between 2003-2008, the federal reserve is more inclined to increase the rate to have room to decrease if a crisis would occur similarly high growth periods tend to have high inflation which is regulated by the Federal Reserve with increasing the Federal Funds rate. Similarly in recessions, the federal reserve is more likely to decrease the rate to stimulate the economy. As the Federal Funds rate is highly influenced by macroeconomic conditions we see a positive relationship with the FFR and MVICPrice. Even though valuations should be changed by the differing interest rates, the overall economic trends seem to have a stronger influence on the prices. This could be due to more optimistic future growth estimations for companies when the economy is growing fast, which translates to higher values, or due to higher demand in the acquisition market as companies are more willing to buy when the economy is booming, which can be seen in the trends in the acquisition numbers.


Another issue with the study has to do with the implementation of the regulation variables. The variables used in the study are factor variables with two levels based on when the regulation in question was enacted. This poses an issue as the variables are simply one point in time and we look at the prices before and after that point. Even though a regulation piece might have an impact on the prices it is more likely that we are only seeing the differences before and after that point in time, which are probably affected by multiple other factors and not only the regulation piece. For example, Regulation FD was enacted in 2000, when the dot com boom was still going, the economy was growing, and the private acquisition market was at high levels. The variable catches the high levels before the regulation and after, which included two recessions and a long stagnation period which all affected the acquisitions market. The coefficient result do not speak fully to the direct impact of the regulation, but rather the overall change in the market before and after the regulation which can be attributed to multiple different factors. 

Another issue with the regulation variable has to do with the number acquisitions before and after the enactment. Three of the four regulation pieces were enacted in the early 2000's and only a small portion of the dataset has acquisitions before the dates and a large portion has acquisitions after the date. This assumes that the regulation will have a lasting and constant impact on the acquisitions from the enactment to 2018, which is probably not the case as companies adapt.

In conclusion the results of this study seem to be inconclusive. They seem to be contrary to what was expected, which is most likely due to the issue that all of these variables tend to be tied to overall macroeconomic environment. Even though statistically speaking we see significant impact from these variables there doesn’t seem to be any actual causal relationship, but rather a correlation to the conditions at that point in time.


# Bibliography
